---
title: "Data structures used in `modstatespace`"
output: rmarkdown::html_vignette
bibliography: references.bib
vignette: >
  %\VignetteIndexEntry{Data structures used in `modstatespace`}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r knitr, include=FALSE}
knitr::opts_chunk$set(
  collapse=TRUE,
  comment="#>",
  fig.align='center', crop=T,
  out.width="99%",
  fig.height = 3.9, fig.width = 8
  )
```

```{r setup}
library(modstatespace)
library(fpp3)
```

`modstatespace` uses `tsibble` from the package `tsibble` (part of the tidyverse) as the data container.
This has the advantage that one can use all the amenities from the package for real 
world data series. The downside is that the `tsibble` data container is somewhat clumsy for multivariate, simulated, balanced time series. Hence we need converter functions to compile data matrices from the `tsibble`. This vignette provides the basis for working with `tsibble` in the context of `modtstatespace`. 

# Working with `tsibble` 
Imagine you have s simple quarterly time series `ts_quart`, or a yearly times series `ts_year`:

```{r}
(ts_quart <- sin(ts(1:20, frequency=4,  start=c(1959, 2))))
(ts_year  <- cos(ts(1:34, frequency=12, start=c(1959, 2))))

qtr <- yearquarter("1959 Q2")

ts_quart <- tsibble(
  time = qtr+(0:19),
  y = sin(1:20),
  index = time
)

ts_quart

month <- yearmonth("1959 Feb")

ts_month <- tsibble(
  time = month + (0:33),
  y = cos(1:34),
  index = time
)

ts_month
```

The data can be plotted easily:

```{r}
autoplot(ts_quart)
```
Note, however, that the intuitive command `plot(ts_quart)` produces a scatter plot rather than a time series plot. 

The plot can be manipulated using the `ggplot2` structure by adding components to the plot, for details see [@Hyndman], Chapter 2.

```{r}
autoplot(ts_month) +
  labs(title = "Monthly data",
       subtitle = "Cosine wave",
       y = "Observations")
```

By using the `tsibble` data structure, we may use the features like seasonal plotting: 

```{r}
gg_season(ts_month)
```


# Inspecting the autocorrelation

The autocorrelation present in the time series can be investigated visually by plotting the time series against its lags: 

```{r}
gg_lag(ts_month, geom = "point")
gg_lag(ts_month, geom = "path")

```

In order to get some more quantitative information, one uses the pair of ACF and PACF: 

```{r}
ACF(ts_month, lag_max = 10)
PACF(ts_month, lag_max = 10)

```
The command generates `tsibbles` containing the ACF and PACF data. These need to be plotted to obtain the familiar autocorrelogramm plots:

```{r}
ACF(ts_month, lag_max = 10) |> autoplot() + labs(title ="ACF")
PACF(ts_month, lag_max = 10) |> autoplot() + labs(title ="PACF")

```


# Working with multivariate data sets 
Next consider you do not have only one observation in each month but multiple: 

```{r}
month <- yearmonth("1959 Feb")

ts_month <- tsibble(
  time = month+(0:33),
  y1 = cos(1:34),
  y2 = rnorm(34,0,1),
  index = time
)

ts_month
```
Now plotting must indicate, which variable is to be plotted: 

```{r}
autoplot(ts_month,y2)
```

Alternatively, we may introduce an additional variable indicating the different variables: 

```{r}
month <- yearmonth("1959 Feb")

t_rep <- c(month+ (0:33),month + (0:33))
obs <- c(cos(1:34),rnorm(34,0,1))
vars <- c(rep("y1",34),rep("y2",34))

ts_month <- tsibble(
  time = t_rep,
  y = obs,
  vars = vars,
  index = time,
  key = vars
)

ts_month
autoplot(ts_month)
```

The additional variable `key` provides the key to the various variables. The plotting command then combines all variables in a single plot. Extracting one coordinate then can be achieved by filtering: 

```{r}
ts_month |> filter( vars == "y2") |> autoplot()
```
This is, how `ACF` and `PACF` react to multivariate data: 

```{r}
ACF(ts_month) |> autoplot()
PACF(ts_month) |> autoplot()
```

# Converting balanced `tsibble` to matrices 
For our procedures we require the data to be stored in a matrix object with rows 
corresponding to time points and columns to variables. The conversion of a `tsibble` 
assumes that the multivariate time series is balanced such that for each time step 
there exists exactly one observation for each coordinate.

```{r}
dat_mat <- tsibble_to_mat(ts_month)
dat_mat
```

This assumes that the data is stored in a variable named `y`. Adding the argument `var_name` one may indicate a different variable name. 

## References 
