---
title: "Estimating a state space system of order n"
output: rmarkdown::html_vignette
bibliography: references.bib
vignette: >
  %\VignetteIndexEntry{Estimating a state space system of order n}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r knitr, include=FALSE}
knitr::opts_chunk$set(
  collapse=TRUE,
  comment="#>",
  fig.align='center', crop=T,
  out.width="99%",
  fig.height = 3.9, fig.width = 8
  )
```

```{r setup}
library(modstatespace)
library(fpp3)
```

In this vignette we discuss the estimation of state space models. To this end we generate a simulated time series:

```{r}
A <- diag(2)*0.5
C <- diag(2)
K <- diag(2)
Sigma <- diag(2)

syst <- StSp(A=A, C=C,K=K,Sigma = Sigma, type = "innov")
y <- sim_syst(syst, T=1000,stationary = TRUE)
autoplot(y) 
```

# Using the CVA subspace approach

Subspace methods are a numerical fast method to estimate state space systems. They rely on regressions and hence are very robust. As a rule-of-thumb one can say that CVA provides good results in situations where autoregressive modelling is attractive.

The estimator requires a number of input arguments:

```{r}
y_mat <- tsibble_to_mat(y)
syst_cva <- cal_CVA(t(y_mat), f=5, p=5, n=2)
syst_cva
```
Here $f$ and $p$ in the literature are suggested to be chosen according to $2\hat k_{AIC}$, where $\hat k_{AIC}$ denotes the AIC estimated lag length in an autoregressive approximation.  

We can see that the estimate is close to the true system. We can compare $-2$ times the log of the Gaussian likelihood at the estimate with the one at the true system:

```{r}
theta_0 <- syst_param(syst)
theta_est <- syst_param(syst_cva)
theta_est-theta_0

cal_crit(as.vector(theta_est),n=2,s=2,t(y_mat))
cal_crit(as.vector(theta_0),n=2,s=2,t(y_mat))
```

So CVA ist better than the true vector in the sense of the Gaussian likelihood.

# Quasi Maximum Likelihood Estimation (qMLE)

Using CVA as a starting value we may optimize the Gaussian likelihood.

```{r}
(theta_hat_pe <- nlm(cal_crit, p=theta_est, n=2, s=2, y=t(y_mat), pe =TRUE , print.level = 0, hessian = TRUE))

cal_crit(as.vector(theta_hat_pe$estimate),n=2,s=2,t(y_mat))

```

The above minimization used the prediction error version of the estimator. We may also derive the actual Gaussian likelihood taking stationary initialization into account:

```{r}
suppressWarnings(theta_hat_pe <- nlm(cal_crit, p=theta_est, n=2, s=2, y=t(y_mat), pe =FALSE , print.level = 0, hessian = TRUE))

cal_crit(as.vector(theta_hat_pe$estimate),n=2,s=2,t(y_mat))

```

This generates an estimated system and the corresponding standard errors: 

```{r}
syste <- param_syst(theta_hat_pe$estimate, n=2, s=2)

theta_sd <- sqrt(diag(solve(theta_hat_pe$hessian)))

summary_qmle( theta_hat_pe, n=2,s=2)
```

The function `est_qml` combines the steps above into a single command: 

```{r}
out_est <- est_qml(t(y_mat), n=2, kmax = 10, pe=TRUE, print.level=0,iterlim = 1000)

summary_qmle( out_est,n=2, s=2)
```



