---
title: "Estimating an autoregression of order p"
output: rmarkdown::html_vignette
bibliography: references.bib
vignette: >
  %\VignetteIndexEntry{Estimating an autoregression of order p}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 72
---

```{r knitr, include=FALSE}
knitr::opts_chunk$set(
  collapse=TRUE,
  comment="#>",
  fig.align='center', crop=T,
  out.width="99%",
  fig.height = 3.9, fig.width = 8
  )
```

```{r setup}
library(modstatespace)
library(fpp3)
```

Autoregressive processes are the workhorse for time series modelling.
They allow to model the temporal correlation in a form similar to
regressions. This includes specification, estimation and prediction.
This package provides functionality to handle autoregressions.

## Definition

A process $(y_t), t \in {\mathbb N}, y_t \in {\mathbb R}^s,$ is called
an autoregressive process, if it is a solution to the equation

$$y_t = A_1 y_{t-1} + \cdots + A_p y_{t-p} + \varepsilon_t, t=1,2,...$$
for a white noise process
$(\varepsilon_t), t \in {\mathbb N} \in {\mathbb R}^s$. In this package
we only consider processes of zero expectation (so far, this is work in
progress).

The name *autoregression* stems from the form of the definition wherein
the process is explained via a regression form with the regressor being
lagged terms of the process itself. The integer $p$ is called the *lag
length* or *order* of the autoregression, if the corresponding matrix
$A_p$ is non-zero.

The autoregressive equation always has a unique solution for given
observations $y_0,...,y_{1-p}$. In the special case $p=1$ we have

$$y_t = A_1 y_{t-1} + \varepsilon_t$$

which can be iterated to get

$$y_t = \varepsilon_t + A_1 \varepsilon_{t-1} + ...+ A_1^{t-1} \varepsilon_1 + A_1^t y_0.$$

It follows that the properties of $y_t$ depend on the properties of the
matrix $A_1$: If all eigenvalues of $A_1$ are smaller than one in
absolute value, then $A_1^t \to 0$ for $t \to \infty$ and the dependence
on the past vanishes as time passes. In this case the autoregression is
*stable*, otherwise it is unstable.

If some eigenvalues of $A_1$ have modulus $1$, however, the observation
$y_0$ exerts a non-diminishing influence for all later time periods.

For a stable autoregression there exists a special solution: Letting
$y_0 = \sum_{j=0}^\infty A_1^j \varepsilon_{-j}$ (thus using the white
noise for $t \in {\mathbb Z}$) we obtain a *stationary* solution that
has (I) constant zero expectation, (II) constant variance solving the
Lyapunov equation

$$P = A_1 P A_1' + \Sigma$$

where $\Sigma \in {\mathbb R}^{s \times s}$ denotes the variance of
$\varepsilon_t$; and (III) constant covariance
$\mbox{Cov}(y_t, y_{t-k}) = A_1^kP$.

For a scalar process thus the correlation between $y_t$ and its lag
$y_{t-k}$ equals $A_1^k$ and the variance equals $\Sigma/(1-A_1^2)$
(demonstrating the importance of the stability condition, which in the
scalar case amounts to $|A_1|<1$).

Similar to differential equations we may rewrite an autoregression of
order $p$ into a larger autoregressive system of order $1$:

$$y_t = A_1 y_{t-1} + \cdots + A_p y_{t-p} + \varepsilon_t \Rightarrow 
Z_t = \begin{pmatrix} y_t \\ y_{t-1} \\ \vdots \\ y_{t-p+1} \end{pmatrix} 
= \underbrace{\begin{pmatrix} A_1 & A_2 & \cdots &\cdots & A_p\\ I_s & 0 & \cdots & \cdots & 0 \\ 0 & \ddots & \ddots & & \vdots \\ \vdots & \ddots & \ddots & \ddots & \vdots\\ 0 & \cdots & 0 & I_s & 0 \end{pmatrix}}_{\Pi}  Z_{t-1} + \underbrace{\begin{pmatrix} I_s \\ 0 \\ \vdots \\ 0 \end{pmatrix}}_{K}  \varepsilon_t. $$

Thus, we obtain $Z_t = \Pi Z_{t-1} + K \varepsilon_t$. The stability
condition now concerns the eigenvalues of the so called *companion*
matrix $\Pi \in {\mathbb R}^{sp \times sp}$.

Using this rewriting the AR(p) case can be seen as a special case of the
AR(1) model.

Checking the stability assumption can be done in two ways: consider the
scalar AR(1) process with $A_1 = 0.9$ (stable case).

```{r}
A <- matrix(0.9,1,1)
(stab_cond <- check_stability_ar(A))
```

The same procedure for $A_1 = 1.5$ detects the instability:

```{r}
A <- matrix(1.1,1,1)
(stab_cond <- check_stability_ar(A))
```

Note that the pole in this case is inside the unit circle.

The procedures also work for multivariate processes of lag order larger
than one:

```{r}
A <- matrix(rnorm(8),2,4)
(stab_cond <- check_stability_ar(A))

A <- matrix(rnorm(8),2,4)*0.1
(stab_cond <- check_stability_ar(A))
```

## Simulating from an autoregression

In this package a time series corresponding to an autoregression can be
simulated using the function `sim_ar`:

```{r}
A <- matrix(0.9,1,1)
y_df <- sim_ar(A,Sigma=1,T=100,stationary =TRUE)
autoplot(y_df)
```

An unstable AR(1) equation leads to an exponentially growing series:

```{r}
A <- matrix(1.1,1,1)
y_df <- sim_ar(A,Sigma=1,T=100,stationary =TRUE)
autoplot(y_df)
```

Also multivariate time series for lags higher than one can be simulated:

```{r}
A <- cbind(matrix(0,2,2),diag(2)*0.9)
y_df <- sim_ar(A,Sigma=diag(2),T=1000,stationary =TRUE)
autoplot(y_df)
```

## Estimation

Since the autoregressive equation has the form of a regression, ordinary
least squares (OLS) is the prime method for estimating the model.
[@SchDei] provide also other estimators, the Yule-Walker estimator being
the prime competitor. In the package only OLS estimation is performed.

For OLS a regression is estimated for each coordinate process
independently. The sample size needs to be reduced for these
regressions, as for the first $p$ observations $y_1,...,y_p$ not all
regressors have been measured. Thus for $y_t$ observed for $t=1,...,T$
only $T-p$ equations can be included in the regression reducing the
effective sample size to $T-p$.

Otherwise regression is straightforward. The procedure `fit_ar` takes as
arguments: + `y` the data set needs to be either of the structure
obtained by simulating as above; or it needs to be a matrix with $s$
rows and $T$ columns.

- `k` the lag order; if $k=-1$ (default) lag length selection using
  `AIC` is done.

-  `kmax` the maximal lag order for`AIC` lag length selection.

If the lag length is selected, then one model for each lag length is
estimated based on the same subset of the data corresponding to the
effective sample size for the largest model.

```{r}
ar2 <- fit_ar(y_df,k=2)
summary(ar2)
```

The summary output first provides the estimate in the form of matrices.
Subsequently the regressions using the base R function `lm` for each
coordinate of the observed process is printed. The numbers are not
exactly identical, as the OLS estimates implemented in `fit_ar` use the
maximum likelihood estimates for the variance, while `lm` uses unbiased
errors.

## References
