---
title: "Simulating from a state space system"
output: rmarkdown::html_vignette
bibliography: references.bib
vignette: >
  %\VignetteIndexEntry{Simulating from a state space system}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r knitr, include=FALSE}
knitr::opts_chunk$set(
  collapse=TRUE,
  comment="#>",
  fig.align='center', crop=T,
  out.width="99%",
  fig.height = 3.9, fig.width = 8
  )
```

```{r setup}
library(modstatespace)
library(fpp3)
```

In this vignette we demonstrate how to generate data from a state space system. In order to to do, we must specify the following objects:

-   the state space system,
-   the variance matrices,
-   the sample size,
-   whether the data should be drawn from the stationary distribution (if this exists).

## Specifying the state space system

A state space system is defined using two equations: 

$$
\begin{array}{ccc}
y_t & = & C x_t + v_t, \\
x_{t+1} & = & A x_t + w_t. 
\end{array} 
$$

Here $y_t$ is the *observed process* of dimension $s$ while $x_t$ denotes the latent (unobserved, that is) *state process*. Both processes are defined for $t=1,...,T$. Thus we assume that the process is observed at discrete, equally spaced time points. This is realistic for many data sets including:

-   economic data measured at quarterly, monthly, weekly, daily frequency
-   panel data sets obtained in consecutive panel waves 
-   sensor data observed with a sampling frequency typically specified in Hertz.

The state space system is defined by providing the matrices $A \in R^{n \times n}$ and $C \in R^{s \times n}$. Additionally the characteristics of the noises $v_t$ and $w_t$ must be specified. $v_t \in R^{s}$ is called measurement or observation noise and arises in the *observation equation*. 
$w_t \in R^{n}$ is called state noise and arises in the *state equation*. 
In this package it is always assumed that the joint noise process $(v_t',w_t')'$ is (weak) *white noise* with zero expectation. This implies that the covariance between noises at different time points is zero at
all lags. 

## The variance matrices 
The variance of the joint noise process is partitioned as follows:

$$
V( \begin{pmatrix} v_t \\ w_t \end{pmatrix}) = \begin{pmatrix} R & S' \\ S & Q \end{pmatrix}. 
$$

There exists two special cases: 

-  $w_t = Kv_t$: We obtain $S= KR, Q= KRK'$. This corresponds to the *innovation representation* of the process which is related to the Wold decomposition for stationary processes. 
- $S=0$ such that measurement and state noise are uncorrelated. 


## Observability, controllability and minimality 
As the state is not observed directly, several representations of the state space system are equivalent. 
It is shown in [@HanDei] that for a stationary process every state space system can be brought into 
innovation form. Thus it is no restriction of generality to only work with the innovation representation. 

But also the innovation representation is not unique. One source of non-uniquesness is the choice of the 
basis of the state space: If the state space coordinate system is changed, the state space system still describes the same input-output map relating the observations $y_t$ to the noise processes $v_t, w_t$.

A second sources of non-uniqueness comes from redundancies that may be present in the state space system. 
Consider the following system: 

$$
\begin{array}{ccc}
y_t & = & \begin{pmatrix} A_1 & 0 \end{pmatrix} x_t + v_t, \\
x_{t+1} & = & \begin{pmatrix} x_{t+1,1}  \\  x_{t+1,2} \end{pmatrix} = \begin{pmatrix} A_{11} & 0 \\ A_{21} & A_{22} \end{pmatrix} x_t + \begin{pmatrix} K_{1}  \\ K_{2}  \end{pmatrix} v_t. 
\end{array} 
$$

It follows that the second part of the state $x_{t,2}$ does not influence the observations $y_t$ and also 
not the first part of the state $x_{t,1}$. Therefore, when only observing $y_t$, the value of $x_{t,2}$ is of no importance and we could instead reduce the model to 

$$
\begin{array}{ccc}
y_t & = & C_1 x_{t,1} + v_t, \\
x_{t+1,1} & = & A_{11}  x_{t,1} + K_1 v_t. 
\end{array} 
$$

This is an example of a *not observable* system. 

Analogously consider the system 

$$
\begin{array}{ccc}
y_t & = & \begin{pmatrix} C_1 & C_2 \end{pmatrix} x_t + v_t, \\
\begin{pmatrix} x_{t+1,1}  \\ x_{t+1,2} \end{pmatrix} & = & \begin{pmatrix} A_{11} & A_{12} \\ 0 & A_{22} \end{pmatrix} x_t + \begin{pmatrix} K_{1}  \\ 0  \end{pmatrix} v_t. 
 \\
y_t & = & C_1 x_{t,1} + v_t + C_2A_{22}^t x_{0,2}, \\
x_{t+1,1} & = & A_{11} x_{t,1} + K_1 v_t + A_{12}A_{22}^t x_{0,2} 
\end{array}
$$
Thus we may use the reduced system $(A_{11},C_1,K_1)$ treating the impact of $x_{0,2}$ as an additional 
deterministic process. The original system is not *controllable*. 

The corresponding concepts can be found in [@HanDei], Chapter 1. We can combine them to the concept of *minimality*: A state space system is called *minimal*, if it is observable and controllable. This is linked to the following three matrices: 

$$
\begin{array}{ccc}
{\mathcal O} = \begin{pmatrix} C \\ CA \\ CA^2 \\ \vdots  \end{pmatrix}, & 
{\mathcal C} = \begin{pmatrix} K & AK & A^2K & \cdots  \end{pmatrix}, &
{\mathcal H} = {\mathcal OC}.
\end{array} 
$$
The matrix ${\mathcal O}$ is called *observability matrix*. The system is observable, iff this matrix has
full column rank $n$. The matrix ${\mathcal C}$ is called *controllability matrix*. The system is 
controllable, if this matrix has full row rank. The matrix ${\mathcal H}$ is called *Hankel matrix*. A 
system is called minimal, if the matrix $A \in R^{n \times n}$ where $n$ equals the rank of the Hankel 
matrix. 

In **modstatespace** a system can be reduced to minimality using the function `model_reduce`, that is implemented only for innovation representations. 

## Stationarity 
For the simulation an initial state needs to be provided. There are three main options in the literature:

-  zero initial state 
-  initial state corresponding to the stationary distribution
-  random initial state.

The first two can be achieved by choosing the Boolean argument 'stationary' in the procedure `sim_syst`. 
The third one often is obtained when simulating an initial 'burn in' period that is not used 
subsequently. 

### Drawing from the stationary distribution of the state

The state process is an autoregressive process of order $1$ and has the Markov property. In this package 
it is always assumed that the noise has expectation zero at all times. 
Therefore, the state also has expectation zero and  its variance evolves as follows:

$$
V(x_{t+1}) = P_{t+1} = A P_t A' + Q.
$$
Consequently stationarity leads to the equation 

$$
P = A P A' + Q.
$$

Such equations are called *Lyapunov equations*. As they are linear in $P$ they can be solved by 
vectorization. Since $Q= Q'$ is symmetric, the solution $P$ is symmetric. Being a variance matrix $P$ 
must also be positive semi-definite. This is equivalent to all eigenvalues of $A$ being smaller than one in modulus.

Such matrices are called *stable*. Under stability then the Lyapunov equation delivers the variance of the state corresponding to the stationary solution of the state equation. 

# Example simulation

The theory above is demonstrated in this section. We follow the steps outlined above. First we specify a system, which we do on purpose non-minimal. 

```{r}
A <- 0.5* diag(3)
C <- matrix(0,2,3)
C[1,1] <- 1
C[2,2] <- 1 
C[2,1] <- 0.5 
Q <- diag(3) 
S <- matrix(0,3,2)
R <- diag(2)

syst <- StSp(A=A,C=C,Q=Q,S=S,R=R,type = 'UC')
syst

```

With this system we can simulate a time series of length $T=1000$ with zero initial value: 


```{r}
set.seed(1)
y <- sim_syst(syst,T= 1000,stationary=FALSE)
autoplot(y)
```

Alternatively the initial state can be drawn from the stationary distribution: 

```{r}
set.seed(1)
yst <- sim_syst(syst,T= 100,stationary=TRUE)
autoplot(yst)
```

The two simulated series can be plotted to compare them:

```{r}
add_st <- function(st){paste(st,"st",sep='')}
yst_2 <- yst |>  mutate(vars =  add_st(vars))
yd <- y |> add_row(yst_2)
yd |> filter(time<10) |> autoplot()

```

We see that they differ only in the first couple of instants. 
As we know that the system is not minimal, we can reduce the order:

```{r}
systr <- model_reduce(syst)

```

Last we can compare the covariances between the original and the reduced system: 

```{r}
syst_v <- list()
syst_v[[1]] <- syst
syst_v[[2]] <- systr
class(syst_v) <- "StSp"
cov_orig <- covariance_function(syst_v,M=20,plots=1)
```


## References 