---
title: "Time Series Indicators"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Time Series Indicators}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r knitr, include=FALSE}
knitr::opts_chunk$set(
  collapse=TRUE,
  comment="#>",
  fig.align='center', crop=T,
  out.width="99%",
  fig.height = 3.9, fig.width = 8
  )
```

```{r setup}
library(modstatespace)
library(fpp3)
```

In this vignette we discuss a number of indicators used to characterise time series. Additionally we also discuss ways to characterise state space systems. 

# Generating a state space system 
In this vignette for purposes of illustrating the various indicators we simulate 
a bivariate time series of length $T=100$: 

```{r}
A <- diag(3)
A[1,1] <- 0.5 
A[2,2] <- -0.5
A[3,3] <- 0.9
C <- matrix(0,2,3)
C[1,1]<-1
C[2,2]<-1
C[1,3] <- 0.2
C[2,3] <- -0.2

set.seed(1)
K <- -matrix(runif(6),3,2)*0.5

(syst <- StSp(A =A,C=C,K=K,Sigma= diag(2),type = "innov"))

```

# Time series generation 
With this system we can generate a time series of length $T=100$. We use the stationary distribution of the three dimensional state to start the time series effectively assuming that the data generating system has been running forever: 

```{r}
set.seed(1)
y <- sim_syst(syst,T= 100,stationary=TRUE)
autoplot(y)
```

# ACF and PACF 
The time series properties can be investigated using the ACF and the PACF plots: 

```{r}
ACF(y) |> autoplot()
PACF(y) |> autoplot()
```

The estimated ACF can be compared to the covariance sequence corresponding to the state space system:

```{r}
cv <- plot_cov(syst, channel_in=1:2, channel_out=1:2, L =20 )
cv
```

# Predictions 
Based on this model we can provide predictions: 

```{r}
pr <- predict(syst,y,T_est_val=50,h=1)
p<-plot_predict(pr)
```

